---
title: "Reproducible Research Course Project (Week 2)"
date: "May 11, 2016"
author: Jeremias Lalis
output: html_document
---

Setting up the defaults:
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, results = "asis")
```

Code for downloading and reading in the dataset:
```{r}
if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")) {
        temp <- tempfile()
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
        unzip(temp)
        unlink(temp)
}
rawData <- read.csv("activity.csv", stringsAsFactors = FALSE)
dim(rawData)
```

Code for preprocessing the data (Removing NAs):
```{r}
noNAData <- rawData[complete.cases(rawData), ]
dim(noNAData)
```

###Q1: What is mean total number of steps taken per day?
####Ans:

Histogram of the total number of steps taken each day.
```{r}
library(ggplot2)
ggplot(noNAData, aes(as.factor(date),steps))+geom_bar(fill="royalblue3", stat="identity")+xlab("Day") + ylab("Steps")+ggtitle("Histogram of the total number of steps taken each day")+ theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}

meanFreq <- as.integer(mean(aggregate(steps~date,noNAData, FUN=sum)$steps))
```
Mean number of steps taken each day is `r meanFreq`.

```{r}
medianFreq <- median(aggregate(steps~date,noNAData, FUN=sum)$steps)
```
Median number of steps taken each day is `r medianFreq`.

###Q2: What is the average daily activity pattern?
####Ans:
```{r}
q2 <- noNAData
q2 <- aggregate(steps~interval, q2, FUN=mean)
plot(q2$steps ~ q2$interval, type="l", xlab="Date", ylab="Average Steps per Day (5-mins interval")
```

Code to look for the maximum number of steps:
```{r maxstep, results="hide"}
maxSteps <- q2[q2$steps==max(q2$steps), ]
q2 <- NULL
```

The 5-minute interval `r maxSteps$interval`, on average across all the days in the dataset, contains the maximum number of steps (`r maxSteps$steps`).

###Q3: Imputing missing values
####Ans:
Code to calculate the total number of missing values in the dataset.
```{r missingval, results="hide"}
numNA <- nrow(rawData) - nrow(noNAData)
```

The total number of rows with NAs in the dataset is `r numNA`.

The simple strategy that I used in imputing the data is that I used the average steps taken for that specific interval in all of the data set and assign it to the missing value(s). The code for this strategy is the following:
```{r}
imputedData <- rawData
#The mean for that 5-minute interval
meanForNA <- aggregate(steps~interval,imputedData, FUN=mean)
for(i in 1:nrow(imputedData)){
  if(is.na(imputedData[i,c("steps")])){
    imputedData[i,c("steps")] <- meanForNA[meanForNA$interval==imputedData[i,c("interval")], c("steps")]
  }
}
```

Code to Show Histogram of the total number of steps taken each day after missing values are imputed.
```{r}
ggplot(imputedData,aes(as.factor(date),steps))+geom_bar(fill="royalblue3", stat="identity")+xlab("Day") + ylab("Steps")+ggtitle("Histogram of the Total Number of Steps Taken Each Day (Imputed Data)")+ theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
newmeanFreq <- as.integer(mean(aggregate(steps~date,imputedData, FUN=sum)$steps))
newmedianFreq <- as.integer(median(aggregate(steps~date,imputedData, FUN=sum)$steps))
```

The new mean of the imputed data is `r newmeanFreq` steps compared to the old mean of `r meanFreq` steps. That creates a difference of `r newmeanFreq - meanFreq` steps on average per day.

The new median of the imputed data is `r newmedianFreq` steps compared to the old median of `r medianFreq` steps. That creates a difference of `r newmedianFreq-medianFreq` steps for the median.

This shows that the overall shape of the distribution has not changed after applying the method in imputing the data.

###Q4: Are there differences in activity patterns between weekdays and weekends?
####Ans:

Code to Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
weekDayEnd <- data.frame("dayType"=character(0))
weekDayEnd <- NULL

#Check wether the day is a weekday or weekend (Saturday)
weekInd <- function(day){
  if (weekdays(day)=="Saturday") {
    out <- "Weekend" 
  } else {
    out <- "Weekday" 
  }
  out
}

for (i in 1:nrow(imputedData)){
  date <- imputedData[i,c("date")]
  newObs <- data.frame("dayType"= weekInd(as.Date(date)))
  weekDayEnd <- rbind(weekDayEnd, newObs)
}
#------------------------------------------------------------

#Add the day type column in the data set
imputedData <- cbind(imputedData, weekDayEnd)
```

Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends.

```{r}
averageWeekDayEnd<-aggregate(steps~interval + dayType, imputedData, FUN=mean)
ggplot(averageWeekDayEnd,aes(interval,steps, color=dayType))+geom_line()+facet_wrap(~dayType, ncol=1)+xlab("5-min Interval")+ylab("Steps")+guides(fill=FALSE)
```